Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics

### 2019-01-11 scagnostics.Rmd
reworked intro

A large number of projection pursuit indexes have been developed, to detect departure from multivariate normal, which includes clusters or outliers or separations between known groups (e.g. @f87, @CBC92, @lckl2005, @AHC02, @CEM:CEM2568, @JS87, @5508437).  Less work has been done on indexes to find nonlinear dependence between variables. ^[\textcolor{red}{The second component in implementing a projection pursuit method is the optimisation algorithm. Several algorithms have been suggested, some tailored to specific applications, see e.g. XXX references?. In this work our focus is on the index functions, and we will only use optimisation algorithms available via the guided tour implementation of tourr.}]

@@

Our motivation is based on applications in physics, to aid the interpretation of model fits on experimental results. Consider a physical model with a set of $p$ free parameters, that cannot be measured directly and are determined by fitting a set of $n$ experimental observations, for which predictions can be calculated when fixing all $p$ parameters. Note here, that while we often have analytic expressions for the predictions, this is not always the case and we may have to rely on numerical computation. Moreover, a single prediction can be a complicated function of all free parameters. In addition, we often deal with large number of observations, $n \sim 100-1000$ as well as sizable parameter spaces $p \sim 10$. The results of a fit are generally interpreted using combinations of variables selected by intuition and prior knowledge. But this begs the question, whether important associations are missed because tools to search are not available.

We draw on developments from the related topic of variable selection. With high-dimensional data, even plotting all pairs can be daunting, so "scagnostics" [@scag] [@WW08] have been developed to find the most interesting pairs of variables. There are 8 scagnostics, vividly named: "Outlying", "Skewed", "Sparse", "Clumpy", "Striated", "Convex", "Skinny" and "Stringy". Our question is whether these can be extended into projection pursuit indexes that can be used to find two-dimensional projections of high-dimensional data with unusual features.

[@Grimm2016] explored the behavior of scagnostics for selecting variables, and proposed two more, based on smoothing splines and distance correlation, that have some nicer properties. In addition the availability of another two indices based on information criteria, maximal and total information coefficient (MIC and TIC) [@Reshef1518], and computationally more efficient versions (MIC_e, TIC_e) [@JMLRv1715308], round out the collection of metrics. ^[It is interesting to note that entropy, closely related to the notion of mutual information, has previously been considered as a projection pursuit index when searching for interesting one-dimensional projections, see discussion in @huber85 and @JS87.]
These indices are all available in R packages [@rref]: scagnostics [@HWscagR], mbgraphic [@mbgraphic] and minerva [@minerva]. The projection pursuit guided tour is available in the R package, tourr [@tourr].

@@

A projection pursuit index (PPI) is a scalar function $f$ defined on an $d$-dimensional data distribution. Typically the definition is such that larger values of $f$ indicate more interesting distribution, and therefore maximizing $f$ over all possible $d$ dimensional projections of a data set with $p>d$ parameters will find the most interesting projections. As most indices characterize deviations from a normal distribution, one would first map the empiric data distribution onto a density for which such deviations can then be defined in different ways, see e.g. @CBC93. When departing from the idea of characterizing differences to the normal distribution it may however be preferred to work directly with the empirical data distribution and instead characterize it by measures based on e.g. the minimal spanning tree or the mutual information.

>>>>

A large number of projection pursuit indexes have been developed, primarily based on departure from normality, which includes clusters or outliers, and also for finding separations between known groups (e.g. @f87, @CBC92, @lckl2005, @AHC02, @CEM:CEM2568, @JS87, @5508437).  Less work has been done on indexes to find nonlinear dependence between variables, when $d>1$.

@@

The motivation is based on physics applications, to aid the interpretation of model fits on experimental results. A physical model can be considered to be a set of $p$ free parameters, that cannot be measured directly and are determined by fitting a set of $q (p<q)$ experimental observations, for which predictions can be made once the $p$ parameters are estimated. Different model fits ($n$) yield the data to be examined using projection pursuit. (Note here, that while we often have analytic expressions for the predictions, this is not always the case and we may have to rely on numerical computation.) A single prediction can be a complicated function of all of the free parameters, and typically $q \in [100,1000]$ and $p \sim 10$. Current practice is to examine pairs of parameters, or combinations produced by intuition or prior knowledge. This begs the question, whether important nonlinear associations are missed because tools to search are not available.

To find appropriate projections pursuit indexes, literature on variable selection was examined. With high-dimensional data, even plotting all pairs of variables can lead to many plots, which is what "scagnostics" (@scag, @WW08) were developed to address by providing metrics from which to select the most interesting variable pairs. There are eight scagnostics, descriptively named: "outlying", "skewed", "sparse", "clumpy", "striated", "convex", "skinny" and "stringy". The question is whether these can be adapted into projection pursuit indexes, to search for unusual features in two-dimensional projections of high-dimensional data. Recent PhD research by @Grimm2016 explored the behavior of scagnostics for selecting variables, and proposed two more that have nicer properties, based on smoothing splines and distance correlation. In addition, two more indexes for measuring dependence have been proposed in the machine learning literature, based on information criteria, maximal and total information coefficient (MIC and TIC) [@Reshef1518], with computationally more efficient versions (MIC_e, TIC_e) [@JMLRv1715308]. These are related to original 1D projection pursuit indexes based on entropy (e.g. @huber85, @JS87). This provides 12 current indexes for measuring dependence between two variables, and each is available in an R [@rref] package: scagnostics [@HWscagR], mbgraphic [@mbgraphic] and minerva [@minerva]. The projection pursuit guided tour is available in the R package, tourr [@tourr] provides optimisation routines, and visualization.

<!--
^[It is interesting to note that entropy, closely related to the notion of mutual information, has previously been considered as a projection pursuit index when searching for interesting one-dimensional projections, see discussion in @huber85 and @JS87.]
These indices are all available in R packages [@rref]: scagnostics [@HWscagR], mbgraphic [@mbgraphic] and minerva [@minerva]. The projection pursuit guided tour is available in the R package, tourr [@tourr].

^[\textcolor{red}{The second component in implementing a projection pursuit method is the optimisation algorithm. Several algorithms have been suggested, some tailored to specific applications, see e.g. XXX references?. In this work our focus is on the index functions, and we will only use optimisation algorithms available via the guided tour implementation of tourr.}]
-->

@@

A projection pursuit index (PPI) is a scalar function $f$ defined on an $d$-dimensional data distribution. Typically the definition is such that larger values of $f$ indicate a more interesting distribution of observations, and therefore maximizing $f$ over all possible $d$ dimensional projections of a data set with $p>d$ parameters will find the most interesting projections. As most indices characterize deviations from a normal distribution, one would first map the empiric data distribution onto a density for which such deviations can then be defined in different ways, see e.g. @CBC93. When departing from the idea of characterizing differences to the normal distribution it may however be preferred to work directly with the empirical data distribution and instead characterize it by measures based on e.g. the minimal spanning tree or the mutual information.
